Efforts to Code Calls created a Positive Impact in Reducing Toronto's Homeless Death Counts

### 2024-01-25 outputs/paper/paper.qmd
Stylistic edits were performed on the paper

All data used in this paper are obtained through OpenDataToronto Portal [@OpenDataToronto]. Two different datasets, Central Intake Call Wrap-Up Codes Data [@Central_Intake_Call] and Deaths of People Experiencing Homelessness[@Homeless_Death_Count], are retrieved to analyze the correlation between different factors of Toronto’s Central Call Line efforts and the monthly cumulative death counts of homeless individuals in Toronto. Data was cleaned and analyzed using the open source statistical programming language R [@citeR] and supporting packages tidyverse [@tidyverse], janitor [@janitor], dplyr [@dplyr], lubridate [@lubridate], zoo [@citeZoo], ggplot2 [@ggplot2], and knitr [@knitr]. Detailed process of data extracting and cleaning can be found in the subsections below.

@@

The data set provides a daily summary of the number of calls received, the number of calls classified into distinct wrap-up codes by the nature of the issue, and a count of calls under each wrap-up code. One of the example wrap-up codes in the original data set was: Code 1A - Referral to a Sleeping/Resting Space. The original data set includes 13 distinct wrap-up codes; only two codes, Code 1A — Referral to a Sleeping/Resting Space and Code 2C — Information - Homelessness & Prevention Services, were chosen for our analysis. Code 1A and Code 2C are best suited as measurements for positive impact the Central Intake Line could provide. Other non-suitable wrap-up codes include: Code 1D - Declined Shelter/Resting Space, Code 4B - Disconnected - No Outcome, etc.

The process of data cleaning is used to rename and exclude unnecessary columns. Variables of relevance: Date, Total calls coded, Code 1A, and Code 2C, were kept in the final dataset (@tbl-monthly-calls).

@@

The original dataset contains three columns: Year of death, Month of death, and Count. After careful inspection of raw data, any rows that containing an “unknown” value are excluded. Although this action created a source of error in future analyses, this is still a necessary step because we are plotting death counts against a timeline.

@@

## Using Data from Different Datasets {#sec-data-diff-datasets}
During the data cleaning process, code chunks are written to ensure that both the Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call] and the Deaths of People Experiencing Homelessness Dataset [@Homeless_Death_Count] are filtered to include only data between November 1st, 2020 to June 30st, 2023.

The decision to only include this time period is reached by taking the common time period between the two datasets. In the latest refresh, the Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call] begins on November 3rd, 2020 and end on December 31st, 2023; whereas the Deaths of People Experiencing Homelessness Dataset [@Homeless_Death_Count] begins on January of 2017 and ends on June 2023.

>>>>

All data used in this paper are obtained through OpenDataToronto Portal [@OpenDataToronto]. Two different datasets: Central Intake Call Wrap-Up Codes Data [@Central_Intake_Call] and Deaths of People Experiencing Homelessness[@Homeless_Death_Count], are retrieved to analyze the correlation between different factors of Toronto’s Central Call Line efforts and the monthly cumulative death counts of homeless individuals in Toronto. Data was cleaned and analyzed using the open source statistical programming language R [@citeR] and supporting packages tidyverse [@tidyverse], janitor [@janitor], dplyr [@dplyr], lubridate [@lubridate], zoo [@citeZoo], ggplot2 [@ggplot2], and knitr [@knitr]. Detailed description of each dataset can be found in the subsections below.

@@

The data set provides a daily summary of the number of calls received, the number of calls classified into distinct wrap-up codes by the nature of the issue, and a count of calls under each wrap-up code. One of the example wrap-up codes in the original data set was: *Code 1A - Referral to a Sleeping/Resting Space*. The original data set includes 13 distinct wrap-up codes; only two codes, *Code 1A — Referral to a Sleeping/Resting Space* and *Code 2C — Information - Homelessness & Prevention Services*, were chosen for our analysis. Code 1A and Code 2C are best suited as measurements for positive impact the Central Intake Line could provide because they provide a count of the number of callers provided with directions or advice. Other non-suitable wrap-up codes include: *Code 1D - Declined Shelter/Resting Space*, *Code 4B - Disconnected - No Outcome*, etc.

@@

The original dataset contains three columns: *Year of death*, *Month of death*, and *Count*. After careful inspection of raw data, any rows that containing an “unknown” value are excluded. Although this action created a source of error in future analyses, this is still a necessary step because we are plotting death counts against a timeline.

@@

## Using Data from Different Datasets {#sec-data-diff-datasets}
During the data cleaning process, code chunks are written to ensure that both the Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call] and the Deaths of People Experiencing Homelessness Dataset [@Homeless_Death_Count] are filtered to include only data between 1st November, 2020 to 30st June, 2023.

The decision to only include this time period is reached by taking the common time period between the two datasets. In the latest refresh, the Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call] begins on 3rd November, 2020 and end on 31st December, 2023; whereas the Deaths of People Experiencing Homelessness Dataset [@Homeless_Death_Count] begins on January of 2017 and ends on June 2023.
